---
title: "Intro to R 2025 day 2"
author: "Matt Cannon"
date: '2025-04-01'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Day 2

## General stuff

## How to make an Rmd file
- Open RStudio
- File -> New File -> R Markdown
- Give it a title and author
- Click OK
- You'll get a template with some example text and code
- Write your text and code in the chunks


## packages
- Packages are collections of functions that do specific things. You're generally going to end up using packages specific to whatever you want to do in R.
- You can install packages with `install.packages("package_name")` and load them with `library(package_name)`.
- Some packages need to be installed using `BiocManager::install("package_name")`.
- If the package is from GitHub, you'll use `devtools::install_github("package_name")`.

- The tidyverse package is actually a bundle of other packages and has a ton of very useful functions.
- I almost consider this package to be a requirement for using R.
```{r}
#install.packages("tidyverse", Ncpus = 8)
library(tidyverse)

install.packages("praise")
library(praise)
praise()
```

### Subsetting a data.frame with [ ]
- A dataframe is a variable type.
- It is a table of data with columns and rows. Most of your data will be in data frames.
- You can subset a dataframe with square brackets `[ ]`.
- You'll put two numbers (or two groups of numbers) in between the brackets with a comma between them.
- The first number is the row(s) you want and the second number is the column(s) you want.
- You can remember rows / columns by remembering "Rats each Cheese".
- You can also select names of dataframes using the column's name or a row's name.
- mtcars is a test dataframe that comes with R.

data_frame[row(s), column(s)]

```{r}
# typing a variable will print it out
mtcars
# first row, first column (just one element)
mtcars[1, 1]
# First row, all columns
mtcars["Mazda RX4", ]
# Everything
mtcars[,]
# First column, all rows
mtcars[, 1]
mtcars$mpg
# First ten rows, all columns
mtcars[c(1:10), ]
```

### Subsetting a vector with [ ]
- A vector is a list of things. It has to all be the *same data type*.
- Vectors have only one dimension.
- You can also subset a vector with square brackets `[ ]`.
- But you only need one number or group of numbers since the data are 1D.

```{r}
# A vector
my_vector <- c(1, 2, 3, 4, 5)

# First element
my_vector[1]
# First three elements
my_vector[1:3]
```

## Functions for today's activity

### mean()
This function calculates the average of a set of numbers. It takes in a vector of numbers and returns a single number.
```{r mean}
one_big_number <- c(1, 2, 4, 5, 10000000000)
big_mean <- mean(one_big_number)
big_mean

small_mean <- mean(c(2, 3, 4, 10))
small_mean
```

Many of these functions can get angry about missing data (`NA`).

Include na.rm = TRUE as an argument to ignore missing data

#### This doesn't return what we want
```{r}
mean(c(1, 3, 4, NA, 34))
```

#### This works
```{r}
mean(c(1, 3, 4, NA, 34), na.rm = TRUE)
```

### median()
This function calculates the median of a set of numbers. It takes in a vector of numbers and returns a single number.

Note here, we're using a variable that we created in an earlier chunk. Variables persist across chunks.
```{r median}
median(one_big_number)

mtcars <- mtcars
median(mtcars$mpg)
```

### min()/max()
These functions calculate the minimum and maximum of a set of numbers. They take in a vector of numbers and return a single number.
```{r minmax}
min(mtcars$mpg)

max(one_big_number)
```

### sum()
This function calculates the sum of a set of numbers. It takes in a vector of numbers and returns a single number.
```{r some}
sum(one_big_number)
sum(c(1, 1, 1, 1, 1, 1))
```

### standard deviation
This function calculates the standard deviation of a set of numbers. It takes in a vector of numbers and returns a single number.
```{r sd}
sd(one_big_number)
mpg_sd <- sd(mtcars$mpg)
mpg_sd
```

### filter()
- Filter is a function inside the dplyr package (which is part of tidyverse) that works on dataframes or tibbles (fancy dataframes).
- It filters rows based on some logical condition.
- It keeps rows where the condition evaluates to TRUE.
- You can use `&` for "and" and `|` for "or".
```{r filter}
library(tidyverse)
mtcars <- mtcars

filter(mtcars, mpg > 30)

# and
filter(mtcars, mpg > 30 & wt < 2)

# or
filter(mtcars, mpg > 30 | wt < 2)
```

### select
Pick (or exclude) columns from a dataframe or tibble

`|>` is a pipe - it passes data from one function to the next
%>% is the same thing and you'll see this more, but it requires the magrittr package or tidyverse
    That way you don't have to nest functions (((((((())))))))

head() returns the fist six elements
```{r select}
select(mtcars, mpg, disp)
head(select(mtcars, -mpg, -disp))
select(mtcars, -mpg, -disp) |> head()
select(mtcars, !mpg, !disp) |> head(n = 4)
```

### hist() - histogram
rnorm() samples numbers from a normal distribution
```{r hist}
hist(rnorm(1000, mean = 200, sd = 20), n = 100)
```

### aov()
```{r anova}
aov(mpg ~ disp + wt, data = mtcars)

anova_summary <- summary(aov(mpg ~ disp + wt, data = mtcars))
anova_summary
```

### table()
Count frequencies or combinations
```{r countStuff}
table(c("cat a", "cat a", "cat a", "cat b", "cat b", "cat b"),
      c(1, 1, 2, 2, 2, 2))

table(mtcars$cyl, mtcars$vs)
```

## Activity

### Read in and check out the Titanic data
```{r}

```

### Make variables that contain the Mean/median/min/max of age/fare
```{r}

```

### Use filter to make two new dataframe variables with just male and female only and then calculate the mean/median values for age and fare
Basically, calculate the mean/median age and fare for men vs women
```{r}

```

### Make a histogram of the age all passengers
```{r}

```

### Make another histogram for only passengers who survived

## If you're super fast

### Use `aov()` to test which factors influenced who survived the titanic disaster
```{r}

```

### Use table() to compare how many women vs men survived
```{r}

```

