---
title: "Intro to R 2023 day 3"
author: "Matt Cannon"
date: '2023-10-25'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
```

# Day 3

## General stuff

### Common error messages
```{r, eval=FALSE}
mean(bob)
#Error in mean(bob) : object 'bob' not found
```
R is looking for a variable named bob, but it doesn't exist
-   Need to make it?
-   Check spelling/capitalization
-   You forgot to put quotes around a character string

```{r, eval=FALSE}
arbleGarble(mtcars)
#Error in arbleGarble(mtcars) : could not find function "arbleGarble"
```
R is trying to use the function `arbleGarble()`, but it doesn't exist
-   Either load in the library that has it
    -   (or use `coolLib::arbleGarble()`)
-   Check spelling/capitalization

```{r, eval=FALSE}
hist(mtcars$mpg, n = 20
# +
((((((1))))))
```
R thinks you have unfinished business here
-   Hit escape to cancel the command
-   Most likely you're missing a parenthesis or a quotation mark somewhere
-   Rainbow brackets!

### googling error messages

### googling "R how do I…"

---

## Functions for today's activity

### cbind(), rbind()
attach two data.frames either side to side or top to bottom
rbind will arrange column names
cbind just puts things together without checking order
  Generally a bad idea to cbind() unless you're very careful
```{r}
efficient <-
  mtcars %>%
  filter(mpg > 30)

gas_guzzlers <-
  mtcars %>%
  filter(mpg < 14.5)

rbind(efficient, gas_guzzlers)

# This is total nonsense, but R does it anyways
test <-
  cbind(efficient, gas_guzzlers)
```

### merge()
merge() will attach two data.frames side to side, but match data in shared columns
-   by default, does an inner join
-   use all, all.x or all.y to specify other join types

https://www.datasciencemadesimple.com/wp-content/uploads/2017/06/merge-in-R-2.jpg?ezimgfmt=rs:535x142/rscb1/ng:webp/ngcb1
```{r}
band_members
band_instruments
merge(band_members, band_instruments)
merge(band_members, band_instruments, all = TRUE)
```

### mutate()
Add a new column or change an existing one
```{r}
km_per_mi <- 1.60934

mtcars %>%
  mutate(kpg = mpg * km_per_mi)

mtcars %>%
  mutate(mpg = mpg * km_per_mi)
```

### summarize() and group_by()
group() silently divides up rows by categories
summarize() data within groups in a dataframe
Makes a new column and drops all columns not in group_by() or created by summarize()
```{r}
mtcars %>%
  group_by(cyl) %>%
  summarize(mean_mpg = mean(mpg))

starwars %>%
  group_by(homeworld, species) %>%
  summarize(mean_height = mean(height),
            max_mass = max(mass))
```

### t()
Rotate a dataframe (or matrix) 90 degrees
Also turns the data into a matrix, so we re-convert it using as.data.frame()
```{r}
rotated_cars <-
  t(mtcars) %>% # could also use |>
  as.data.frame()
```

### pivot_longer()
wide data vs long data
"tidy" data - https://www.youtube.com/watch?v=K-ss_ag2k9E explanation starts ~ 6 min in
https://epirhandbook.com/en/images/pivoting/pivot_longer_new.png
https://fromthebottomoftheheap.net/2019/10/25/pivoting-tidily/
```{r}
# wide data
relig_income

# long data
long_data <-
  relig_income %>%
  pivot_longer(!religion,
               names_to = "income",
               values_to = "count")

# Make it wider
wide_again <-
  long_data %>%
  pivot_wider(names_from = "income",
              values_from = "count")
```

### Write a table out to a text file
```{r}
write.table(wide_again,
            file = "exampleOutput.txt",
            quote = FALSE,
            sep = "\t",
            row.names = FALSE)

write_tsv(wide_again,
          file = "exampleOutputAlso.txt")
```

## Activity

### Read in patientGroups.txt and exercise.txt
patientGroups.txt is:

-   patient #
-   treatment groups
-   final blood glucose measurements

exercise.txt is:

-   patient #
-   how many minutes each patient exercised across five days

```{r}

```


### Combine the datasets into a single data.frame
```{r}

```

### Pivot the data to long form
Columns should be patient, day, exercise_min, glucose
```{r}

```

### Save the pivoted data.frame to a text file with write.table() or write_tsv()
```{r}

```

### Make a new column where glucose is multiplied by 1000
```{r}

```

### Calculate the average daily minutes of exercise per patient
```{r}

```

## If you're super fast

### Plot glucose levels for each group (treated/control)
```{r}

```

### Test if glucose levels are statistically different between groups
```{r}

```

### Plot daily exercise minutes per group
```{r}

```

### Test if exercise minutes is statistically different between groups
```{r}

```

