---
title: "Intro to R 2025 day 4"
author: "Matt Cannon"
date: '2025-04-03'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
```

# Day 4

## General stuff

### What is a loop?
- A way to do something over and over again
    - "While I can, do push ups"
- Sometimes it's iterating over a group of things
    - "For each of my shoes, tie the laces"
- Sometimes it's just doing something until a condition is met
    - "If I'm hungry, keep eating ice cream"
- Can also be instructions to do things conditionally
    - "If I have ice cream, eat ice cream"
- Loops are bounded by curly braces { } so you know where they start and end

### Why use a loop?
- Avoid copying and pasting code, which is error prone and time consuming
- Easier to edit afterwards if you want to change something
- Sometimes you have to do a loop
    - Impractical to write the same code 10,000 times
    - You don't know how many times you need to do something

```{r dont_do_this}
plot(1)
plot(2)
plot(2)
plot(3)
plot(4)
plot(5)
plot(6)
```

## Functions for today's activity

### Lists
- A list is like a vector (1 dimensional)
- Each element can be a different data type
- Each element can have a name
- To make a list, put your stuff inside a `list()` function
- To access elements of a list
    - Use `$` and the name of the part if it has a name
    - Use `[[ ]]` and the index (number) of the part if it doesn't have a name
    - **Difference between `[]` and `[[]]`**
        - `[[]]` returns an element of the list
        - `[]` returns a shorter portion of the list

```{r}
bunch_o_stuff <-
    list(
        first = 1,
        some_data = mtcars,
        small_matrix = matrix(c(1, 2, 1, 2), nrow = 2),
        dogs = c("fido", "toby")
    )

bunch_o_stuff$first
bunch_o_stuff$small_matrix
names(bunch_o_stuff)

# What is the difference between this:
bunch_o_stuff[[1]]
# And this?
bunch_o_stuff[1]

bunch_o_stuff[3:4]
```

### if () {}
- "if" loops are used to perform some action if a condition is TRUE
- The condition is inside the parentheses
    - This needs to evaluate down to TRUE or FALSE
- The code to execute is inside the curly braces
- If the condition is FALSE, nothing happens

if ( test condition ) {
  do this stuff if test condition == TRUE
}
```{r}
bobs_age <- 85

bobs_age > 80

if (bobs_age > 80) {
    print("Bob is probably losing hair.")
}
```

#### When to use an if () loop?
- If you need to chose when to perform some action based on some condition
    - The action will not always be performed

if (the day starts with "F") {
    go to the pub
}

if (length(Cells(single_cell_data)) > 1000) {
    DimPlot(single_cell_data)
}

### if/else
- You can give alternatives to perform if the condition is FALSE
- "else" and "else if" () will give alternative conditions to consider
    - "else if()" will take a second condition to check for
    - **The conditions are considered in order**
    - The second condition is only checked if the first is FALSE

if (I have my lunchbox) {
  eat my lunch
} else if (I have money) {
  buy lunch
} else {
  go hungry
}

```{r}
darrins_age <- 15

darrins_age > 80

if (darrins_age > 80) {
    print("Darrin is probably losing hair.")
} else {
    print("Darrin has a full head of luscious hair.")
}

## Can test multiple conditions with else if ()
if (darrins_age > 80) {
    print("Darrin is probably losing hair.")
    print("other stuff")
} else if (darrins_age > 50) {
    print("Darrin might be losing some hair.")
} else {
    print("Darrin has lots of hair.")
}
```

### while () {}
- While the condition is TRUE, do the stuff inside the curly braces
- Use this when you don't know how many times you need to do something
    - In the example with ice cream, if I specified how many bites to take, I may guess the wrong number

while (I have ice cream) {
  eat ice cream
}

```{r}
bettys_age <- 1

while (bettys_age < 100) {
    bettys_age <- bettys_age + 1
    print(paste("Betty ages by a year and is now", bettys_age))
    Sys.sleep(0.3) # This is just here so you can watch the loop happen slowly
}
print("Betty died. :'-(")

rand_num <- rnorm(1, sd = 0.5)
rand_num

tries <- 1
while (rand_num < 0.999) {
    print(
        paste(
            rand_num,
            "I tried to find a number greater than 0.999", tries, "times!"
        )
    )

    rand_num <- rnorm(1, sd = 0.5)
    tries <- tries + 1
}
```

#### Example of an infinite loop
- An infinite loop occurs when the tested condition can never evaluate to FALSE
- This will keep going until you either stop it or the heat death of the universe

while (I'm not a billionaire) {
    Give away all my money
}

```{r eval=FALSE}
bettys_age <- 1
while (bettys_age > 0) {
    bettys_age <- bettys_age + 1
    if (bettys_age > 1000) {
        print("Holy moly!")
    }
    print(paste("Betty is a vampire and can never die. She is now", bettys_age))
}
```

### for () {}
- A for loop is used to iterate over a group of things
- The group of things is defined in the parentheses
- The code to execute is inside the curly braces
- You define a variable name in the parentheses that you will use in the loop
- The variable name will be the current item in the group of things
    - Every iteration of the for loop will use the next item in your group
- Use a for loop when you're doing something to each item in a group of things

for (my_new_variable_name in "list/vector of things") {
    my_new_variable_name now holds one of the things in the list/vector
    I can now do stuff using this iteration of my_new_variable_name
}

for (this_kid in my_kids) {
    feed(this_kid)
}

```{r}
list_of_friends <- list("Bob", "Jack", "Jill")

for (friend in list_of_friends) {
    print(paste("I gave", friend, "a present."))
}

print(friend)

for (i in 1:10) {
    print(paste("I have", i, "ice cream cones."))
}
```

### as.character(), etc…
- Data come in defined types in R
    - numeric is numbers with decimals     3.14
    - integer is whole numbers             3
    - character is text                    "hello"
    - logical is TRUE or FALSE             TRUE
    - Date is a date                       "1983-02-08"
    - factor is a categorical variable     "red", "blue", "green"
        - Similar to characters, but treated like categories
- Sometimes you need to convert data from one type to another
- If you have one thing and you want it to be something else, use as.the_thing_I_want_here()

```{r}
as.character(1)
as.numeric("1")
as.logical(c(1, 0, 2, -1337)) # zero is FALSE, everything ele is TRUE
as.integer(1.2)
as.Date("1983-02-08")
```

#### Beware of coercion!
- R will try to convert data types to make things work
- Sometimes this is helpful, sometimes it's not
- In this case it creates nonsense
- R will sometimes do this silently inside of a function if you give it the wrong data type
    - This can lead to confusing errors
    - Try to keep data types in mind

```{r}
my_numbers <- factor(c("14", "36", "44", "912"))
my_numbers
as.numeric(my_numbers)
```

### colnames()/rownames()
- Get row or column names for a data.frame
- Returns a vector of characters

```{r}
rownames(mtcars)
colnames(mtcars)
```


## Activity

- These activities will be a bit of synthesis of stuff from previous days
- use `starwars <- starwars` to get pre-loaded practice data
- Get rid of the "films", "vehicles" and "starships" columns using `select()` and save it as a new variable
- Write a for loop to loop over the column names in your new variable
    - Inside the loop, if the column contains numeric data, make a histogram of it
    - Otherwise, use `table()` to get counts of categories
        - Look at the help documentation for `table()` to see how to use it
- If you're super fast:
    - Help someone who is struggling (loops can be confusing)
    - Write a while loop to count the sum of all numbers whose squared value is less than 1,234,567
    - You should get 190,519,459,686

### Get the starwars data and look at it
library(tidyverse) # load the tidyverse library
starwars <- starwars # to get data


### Write a for loop using over the “starwars” column names
Inside the loop, if the column contains numeric data, make a histogram of it
Otherwise, use table() to get counts

- Useful ideas/functions here:
    - is.numeric()
    - this_column_name <- "height"
    - starwars[[this_column_name]] # same as starwars$height to select a column

## If you're super fast:
Help someone who is struggling (loops can be confusing)

### Write a while loop to count the sum of all numbers whose squared value is less than 1,234,567

1^2 < 1,234,567   # add 1 to sum total
2^2 < 1,234,567   # add 2 to sum total
3^2 < 1,234,567   # add 3 to sum total
.
.
.
X^2 > 1,234,567   # stop here
