---
title: "Intro to R 2025 day 4"
author: "Matt Cannon"
date: '2025-04-03'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
```

# Day 4

## General stuff

### What is a loop?
A loop is instructions for R to do things either conditionally or repeatedly
Loops are bounded by curly braces { }

### Why use a loop?
Avoid copying and pasting code, which is error prone and time consuming
Easier to edit afterwards if you want to change something
```{r dont_do_this}
plot(1)
plot(2)
plot(2)
plot(3)
plot(4)
plot(5)
plot(6)
```

## Functions for today's activity

### Lists
List of data
Each bit can be named
Can have diverse data types
```{r}
bunch_o_stuff <-
  list(first = 1,
       some_data = mtcars,
       small_matrix = matrix(c(1, 2, 1, 2), nrow = 2),
       dogs = c("fido", "toby"))

bunch_o_stuff$first
bunch_o_stuff$small_matrix
names(bunch_o_stuff)
```

### if () {}
if ( test condition ) {
  do this stuff if test condition == TRUE
}
```{r}
bobs_age <- 85

bobs_age > 80

if (bobs_age > 80) {
  print("Bob is probably losing hair.")
}
```

#### When to use an if () loop
If you need to chose when to perform some action based on some condition
if (the day starts with "F") {
    go to the pub
}

### if/else
Give alternatives to perform if the condition is FALSE

else if () will give alternative conditions to consider

if (I have my lunchbox) {
  eat my lunch
} else if (I have money) {
  buy lunch
} else {
  go hungry
}
```{r}
darrins_age <- 15

darrins_age > 80

if (darrins_age > 80) {
  print("Darrin is probably losing hair.")
} else {
  print("Darrin has a full head of luscious hair.")
}

## Can test multiple conditions with else if ()
if (darrins_age > 80) {
  print("Darrin is probably losing hair.")
  print("other stuff")
} else if (darrins_age > 50) {
  print("Darrin might be losing some hair.")
} else {
  print("Darrin has lots of hair.")
}
```

### while () {}
While the condition is TRUE, do the stuff inside the curly braces

Use this when you don't know how many times you need to do something

while (I have ice cream) {
  eat ice cream
}
```{r}
bettys_age <- 1

while (bettys_age < 100) {
  bettys_age <- bettys_age + 1
  print(paste("Betty ages by a year and is now", bettys_age))
}
print("Betty died. :'-(")

rand_num <- rnorm(1, sd = 0.5)
tries <- 1
while (rand_num < 0.999) {
  print(paste(rand_num, "I tried to find a number greater than 0.999", tries, "times!"))
  rand_num <- rnorm(1, sd = 0.5)
  tries <- tries + 1
}
```

#### Example of an infinite loop
An infinite loop occurs when the tested condition can never evaluate to FALSE

while (I'm not a billionare) {
    spend all my money on oreos
}

```{r}
bettys_age <- 1
while (bettys_age > 0) {
  bettys_age <- bettys_age + 1
  if (bettys_age > 1000) {
    print("holy moly")
  }
  print(paste("Betty is a vampire and can never die. She is now", bettys_age))
}
```

### for () {}
for (variable_name in list/vector of things) {
  do this using this iteration of variable_name
}

for (each of my kids) {
    feed them dinner tonight
}

Use a for loop when you're doing something to each item in a group of things

```{r}
list_of_friends <- c("Bob", "Jack", "Jill")
#list_of_friends <- list(friend1 = "bob", friend2 = "mary")

for (friend in list_of_friends) {
  print(paste("I gave", friend, "a present."))
}

print(current_friend)
```

### as.character(), etc…

Data come in defined types in R

if you have one thing and you want it to be something else, use as.the_thing_I_want_here()
```{r}
as.character(1)
as.numeric("1")
as.logical(c(1, 0, 2))
as.integer(1.2)
as.Date("1983-02-08")
```

#### Beware of coercion!
```{r}
my_numbers <- factor(c("14", "36", "44", "912"))
my_numbers
as.numeric(my_numbers)
```

### colnames()/rownames()
Get row or column names for a data.frame
```{r}
rownames(mtcars)
colnames(mtcars)
```


## Activity

### Get the starwars data and look at it
library(tidyverse) # load the tidyverse library
starwars <- starwars # to get data


### Write a for loop using over the “starwars” column names
Inside the loop, if the column contains numeric data, make a histogram of it
Otherwise, use table() to get counts

is.numeric()
column_name <- "height"
starwars[[column_name]] # same as starwars$height

## If you're super fast:
### Write a while loop to count the sum of all numbers whose squared value is less than 1,234,567

1^2 < 1,234,567   # add 1 to sum total
2^2 < 1,234,567   # add 2 to sum total
3^2 < 1,234,567   # add 3 to sum total
.
.
.
X^2 > 1,234,567   # stop here

Help someone who is struggling (loops can be confusing)

