---
title: "Intro to R 2025 day 5"
author: "Matt Cannon"
date: '2025-04-04'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries
```{r}
library(tidyverse)
```

# Day 5

## General stuff

### geoms as layers
https://ggplot2.tidyverse.org/reference/

## Functions for today's activity, also for reference

### ggplot
ggplot(data_frame_here, aes(x = x_here, y = y_here, other_stuff)) +
layer_1 +
layer_2 +
layer_3 +
. +
. +
. +
modify_the_plot_somehow

I'm not going over all the things in this document, just the ones marked with lots of pound signs ##############################################################

#### geom_point() ##############################################################
Add a scatterplot layer to the plot
```{r}
ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point()

# Set the overall look of the plot - I like the white background better than gray
theme_set(theme_bw())
theme_update(plot.title = element_text(hjust = 0.5))

ggplot(mtcars,
       aes(x = hp,
           y = carb)) +
  geom_point() +
  theme_bw()
```

#### aes(fill/color/shape/size) ################################################
Change colors/shapes/etc

If you put the argument outside of aes()

-   applies to everything in that layer
-   does not show up in the legend

If you put it inside aes()

-   Applies to parts of the layer
-   Shows up in legend

If it is in the aes() call inside of ggplot() `ggplot(aes(...))`

-   It applies to all layers
-   Unless that layer overwrites it

```{r}
mtcars %>%
  ggplot(aes(x = as.factor(cyl),
             y = hp)) +
  geom_boxplot(color = "red",
               fill = "blue",
               size = 4)

ggplot(mtcars,
       aes(x = hp,
           y = carb,
           color = as.factor(vs),
           shape = as.factor(gear),
           size = disp)) +
  geom_point()
```

#### geom_histogram()
Better looking than hist()
Specify how many bars to show with `bins`
```{r}
hist(storms$pressure, n = 200)

ggplot(storms,
       aes(x = pressure)) +
  geom_histogram(bins = 200)
```

#### geom_density()
Smooth your data out
This can actually hide some of the nuance in your data
```{r}
ggplot(storms,
       aes(x = pressure,
           fill = category)) +
  # alpha sets transparency -- 0 is clear, 1 is opaque
  geom_density(alpha = 0.1, adjust = 0.5) +
  # The brewer color palettes are pretty, and "Set2" is color-blind friendly
  scale_color_brewer(palette = "Set2")
```

#### geom_col() ################################################################
Add a barplot layer to the plot
```{r}
mtcars %>%
  rownames_to_column("car_make") %>%
  mutate(car_make = str_remove(car_make," .+")) %>%
  ggplot(aes(x = car_make, y = hp)) +
  geom_col(color = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("this is a title") +
  theme_update(plot.title = element_text(hjust = 0.5))

mtcars %>%
  rownames_to_column("car_make") %>%
  ggplot(aes(y = car_make, x = hp)) +
  geom_col()
```

#### geom_errorbar() ###########################################################
Add error bars as a layer to plot
```{r}
summarized_hp <-
  mtcars %>%
  group_by(cyl) %>%
  summarize(mean_hp = mean(hp),
            sd_hp = sd(hp))

summarized_hp

ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp,
                    xmin = 2,
                    xmax = 2))

# This has error bars in the back since that layer was added first!
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp),
                color = "red") +
  geom_bar(stat = "identity") +
  geom_point(color = "blue")
```

#### geom_jitter()
Drop-in replacement for geom_point()
Adds noise in x and y directions by default so you can see individual points
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp)) +
  geom_jitter(data = mtcars,
              aes(x = as.factor(cyl),
                  y = hp),
              color = "red", height = 0)
```

#### ggbeeswarm package
Adds noise in x direction and pulls things to the center to make it look nicer than jitter
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl), y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp)) +
  ggbeeswarm::geom_beeswarm(data = mtcars,
                            aes(x = as.factor(cyl),
                                y = hp),
                            color = "red")

matt_swarm <- function(...) {
  ggbeeswarm::geom_beeswarm(...)
}
```

#### geom_boxplot() ############################################################
Boxplots!
```{r}
mtcars %>%
  ggplot(aes(x = as.factor(cyl), y = hp)) +
  #geom_boxplot() +
  geom_violin(adjust = 1) +
  ggbeeswarm::geom_beeswarm()


```

#### geom_smooth
Add a regression line to your plot
Can specify the type of regression with `method`
Can remove gray confidence interval with `se = FALSE`
```{r}
ggplot(mtcars,
       aes(x = mpg,
           y = disp)) +
  geom_point() +
  geom_smooth()
```

#### Use multiple datasets in the same plot
Contents of the columns used as axes have to match
Each geom using it's own data needs to have "data = " in it
```{r}
ggplot(summarized_hp,
       aes(x = as.factor(cyl),
           y = mean_hp)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_hp - sd_hp,
                    ymax = mean_hp + sd_hp),
                width = 0.5) +
  geom_point(data = mtcars,
             aes(x = as.factor(cyl),
                 y = hp),
             color = "red")
```

#### labs()
Add labels
```{r}
ggplot(mtcars) +
  geom_point(aes(x = hp, y = disp)) +
  labs(x = "X label goes here",
       y = "Y label goes here",
       title = "This is a title",
       subtitle = "Subtitle!",
       caption = "Cannon et al, 2034!",
       tag = "A")
```

#### facet_wrap() ##############################################################
```{r}
dplyr::storms

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category + status)

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category, scales = "free_y")

ggplot(dplyr::storms,
       aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category,
             nrow = 1,
             scales = "free_y")
```

#### theme()
A good guide to what theme elements are what:
https://henrywang.nl/ggplot2-theme-elements-demonstration/

You can change just about any aspect of how the plot looks with theme()
You'll have to look up how to change each specific bit
```{r}
ggplot(storms,
       aes(x = pressure,
           fill = category)) +
  # alpha sets transparency - 0 is clear, 1 is opaque
  geom_density(alpha = 0.5) +
  # The brewer color palettes are pretty, and "Set2" is color-blind friendly
  scale_color_brewer(palette = "Set2") +
  theme(legend.position = c(0.2, 0.9),
        legend.direction = "horizontal",
        legend.key.width = unit(2, "cm"),
        legend.title = element_text(face = "bold"),
        strip.background = element_rect(color = "white", fill = "white"),
        strip.text.x = element_text(size = 30, face = "bold"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

### ggsave()
```{r}
ggsave("examplePlot.png",
       width = 10,
       height = 8)

plot_in_variable <-
  ggplot(dplyr::storms,
         aes(x = pressure)) +
  geom_histogram(bins = 100) +
  facet_wrap(~ category)

# ggsave will get the plot size from how big your "plot" windows is in the right pane if you don't specify it!!!!!
ggsave("anotherExamplePlot.png",
       plot = plot_in_variable)
```

### heatmap() ##################################################################
If you like heatmaps also check out the pheatmap package
```{r}
heatmap(as.matrix(mtcars))

# Scaling the categories instead
heatmap(as.matrix(mtcars), scale = "column")
pheatmap::pheatmap(as.matrix(mtcars))
```

### PCA
```{r}
mtcars <- mtcars

pca_raw <- prcomp(mtcars, scale = TRUE)

?prcomp

# The output of prcomp is a list
# The element named "x" contains the actual principle components
# Though, it needs to be converted from a matrix to a dataframe for ggplot
pc_values <- pca_raw$x %>%
  as.data.frame() %>%
  rownames_to_column("car_make")

ggplot(pc_values, aes(x = PC1, y = PC2)) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = car_make))
```

## Activity

## Recreate plots from paper

Data is from https://doi.org/10.1371/journal.pbio.2005756

Najafov A, Zervantonakis IK, Mookhtiar AK, Greninger P, March RJ, Egan RK, et al. (2018) BRAF and AXL oncogenes drive RIPK3 expression loss in cancer. PLoS Biol 16(8): e2005756. https://doi.org/10.1371/journal.pbio.2005756

I downloaded their figure data and made it easier to import:
-   Fig1A_partial.txt
-   Fig1C.txt
-   Fig2F.txt
-   Fig3F.txt

#### Figure 1A – geom_boxplot()
```{r}

```

#### Figure 1C – heatmap()
```{r}

```

#### Figure 2F – geom_point()
```{r}

```

#### Figure 3F – geom_col
use geom_errorbars()
and facet_wrap() to split the plots by cell_line
```{r}

```

## If you’re super fast:
Keep going, do the rest 🙃

Download the data from the supplemental data here:

https://doi.org/10.1371/journal.pbio.2005756.s001

