## Activity

- These activities will be a bit of synthesis of stuff from previous days
- use `starwars <- starwars` to get pre-loaded practice data
- Get rid of the "films", "vehicles" and "starships" columns using `select()` and save it as a new variable
- Write a for loop to loop over the column names in your new variable
    - Inside the loop, if the column contains numeric data, make a histogram of it
    - Otherwise, use `table()` to get counts of categories
        - Look at the help documentation for `table()` to see how to use it
- If you're super fast:
    - Help someone who is struggling (loops can be confusing)
    - Write a while loop to count the sum of all integers whose squared value is less than 1,234,567
    - You should get 617,716

### Get the starwars data and look at it
library(tidyverse) # load the tidyverse library
starwars <- starwars # to get data

```{r}
library(tidyverse)

starwars <- starwars
starwars <- starwars %>%
    select(-films, -vehicles, -starships)
```
### Write a for loop using over the “starwars” column names
Inside the loop, if the column contains numeric data, make a histogram of it
Otherwise, use table() to get counts

- Useful ideas/functions here:
    - is.numeric()
    - this_column_name <- "height"
    - starwars[[this_column_name]] # same as starwars$height to select a column

```{r}
for (col in colnames(starwars)) {
    if (is.numeric(starwars[[col]])) {
        hist(starwars[[col]], main = col, xlab = col)
    } else {
        print(table(starwars[[col]]))
    }
    Sys.sleep(1)
}
```


## If you're super fast:
Help someone who is struggling (loops can be confusing)

### Write a while loop to count the sum of all numbers whose squared value is less than 1,234,567

```{r}
sum <- 0
i <- 1
while (i^2 < 1234567) {
    sum <- sum + i
    i <- i + 1
}
print(sum)
```
1^2 < 1,234,567   # add 1 to sum total
2^2 < 1,234,567   # add 2 to sum total
3^2 < 1,234,567   # add 3 to sum total
.
.
.
X^2 > 1,234,567   # stop here
