## Activity
- Use titanic.csv
    - This file is located in the `materials` folder.
    - This is real data on titanic passengers
    - Survived: 1 = yes, 0 = no
- Explore the data
- Make variables that contain average, median, max and min values for age and fare
    - Do the same, but split the data by male/female
- Plot a histogram of age for all the data
    - Plot only the survivors as a second plot
- If you're super fast:
    - Help someone around you who may be struggling
    - Use table() to compare how many men/women survived
    - Use aov() to see which factors influenced survival most

### Read in and check out the Titanic data
```{r}
titanic_data <- read.csv("materials/titanic.csv")
```

### Make variables that contain the Mean/median/min/max of age/fare
```{r}
mean_age <- mean(titanic_data$Age)

median_age <- median(titanic_data$Age)

min_age <- min(titanic_data$Age)

max_age <- max(titanic_data$Age)


mean_fare <- mean(titanic_data$Fare)

median_fare <- median(titanic_data$Fare)

min_fare <- min(titanic_data$Fare)

max_fare <- max(titanic_data$Fare)
```

### Use filter to make two new dataframe variables with just male and female only and then calculate the mean/median values for age and fare
Basically, calculate the mean/median age and fare for men vs women
```{r}
male_data <- filter(titanic_data, Sex == "male")

mean_male_age <- mean(male_data$Age)
mean_male_fare <- mean(male_data$Fare)

female_data <- filter(titanic_data, Sex == "female")

mean_female_age <- mean(female_data$Age)
mean_female_fare <- mean(female_data$Fare)
```

### Make a histogram of the age all passengers
```{r}
hist(titanic_data$Age, n = 100)
```

### Make another histogram for only passengers who survived

## If you're super fast

### Use `aov()` to test which factors influenced who survived the titanic disaster
```{r}
aov(Survived ~ Sex + Age + Fare, data = titanic_data) %>%
    summary()
```

### Use table() to compare how many women vs men survived
```{r}
table(titanic_data$Sex, titanic_data$Survived)

table(titanic_data$Sex, titanic_data$Survived) %>%
    prop.table(margin = 1)
```
