## Activity
- Use patientGroups.txt and exercise.txt from the materials folder
    - patientGroups is patient # and who received treatment
    - Exercise is the how many minutes each patient exercised across five days
- Combine the datasets into a single data.frame
- Pivot the data to long form
    - Columns: patient, day, exercise_min, glucose, trt_group
- Save the pivoted data.frame to a text file with write.table()
- Make a new column by multiplying glucose by 1000
- Calculate the average daily minutes of exercise per patient
- If you're super fast:
    - Help someone who is struggling
    - Plot glucose levels for each group (treated/control)
    - Test if they're statistically different
    - Plot daily exercise minutes per group
    - Test if exercise minutes was statistically different between groups

### Read in patientGroups.txt and exercise.txt
patientGroups.txt is:

-   patient #
-   treatment groups
-   final blood glucose measurements

exercise.txt is:

-   patient #
-   how many minutes each patient exercised across five days

```{r}
patient_groups <- read.table("materials/patientGroups.txt", header = TRUE, sep = "\t")

exercise_data <- read_tsv("materials/exercise.txt")
```

### Combine the datasets into a single data.frame
```{r}
combined_data <- left_join(patient_groups, exercise_data, by = "patient")
```

### Pivot the data to long form
Columns should be patient, day, exercise_min, glucose
```{r}
long_combined_data <-
    combined_data %>%
    pivot_longer(
        cols = starts_with("day"),
        names_to = "day",
        values_to = "exercise_min"
    )

# could also do
combined_data %>%
    pivot_longer(
        cols = c(day_1, day_2, day_3, day_4, day_5),
        names_to = "day",
        values_to = "exercise_min"
    )

# Or
combined_data %>%
    pivot_longer(
        cols = c(-patient, -trt_group, -glucose),
        names_to = "day",
        values_to = "exercise_min"
    )
```

### Save the pivoted data.frame to a text file with write.table() or write_tsv()
```{r}
write_tsv(long_combined_data, "long_combined_data.txt")
```

### Make a new column where glucose is multiplied by 1000
```{r}
long_combined_data <-
    long_combined_data %>%
    mutate(glucose = glucose * 1000)
```

### Calculate the average daily minutes of exercise per patient
```{r}
ave_exercise <-
    long_combined_data %>%
    group_by(patient) %>%
    summarise(ave_exercise = mean(exercise_min, na.rm = TRUE))
```

## If you're super fast

### Plot glucose levels for each group (treated/control)
```{r}
plot(
    as.factor(long_combined_data$trt_group),
    long_combined_data$glucose,
    main = "Glucose Levels by Treatment Group"
)
```

### Test if glucose levels are statistically different between groups
```{r}
t.test(
    glucose ~ trt_group,
    data = long_combined_data
)
```

### Plot daily exercise minutes per group
```{r}
plot(
    as.factor(long_combined_data$trt_group),
    long_combined_data$exercise_min,
    main = "Daily Exercise Minutes by Treatment Group"
)
```

### Test if exercise minutes is statistically different between groups
```{r}
t.test(
    exercise_min ~ trt_group,
    data = long_combined_data
)
```
