## Activity

- Read in “westernQuant.txt” using either read.delim() or the GUI
    - The file can be found in the materials folder. Make sure you tell R where to find it.
    - This is fake data of quantified western blot band intensities
- Use summary() to look at the data
- Use t.test() to test if intensity varies between groups
- Make two plot()s of the two groups
    - Make x-axis group_name for a first plot
    - Make x-axis group_number for a second plot
- Extra task if you're super fast and got done quick
    - Look up wilcox.test() and use it to compare between groups

If you have trouble, look at how I used these functions above as reference, or check out the documentation.

### Read in data
Use read.delim()
```{r readData}
western_quant <- read.delim("materials/westernQuant.txt", header = TRUE, sep = "\t")
```

### Summarize the data.frame
Use summary()
```{r summaryWestern}
summary(western_quant)
```

### Test for group differences
Use t.test() function
```{r testDiff}
t_test_out <- t.test(western_quant$intensity ~ western_quant$group_name)

t_test_out
```
